Understanding the nanoscale structure of hexagonal phase lyotropic liquid crystal membranes

### 2018-02-06 Ben_Manuscripts/structure_paper/Draft.tex

  \begin{abstract}

  Nanostructured porous membranes made from the cross-linked hexagonal phase of
  self-assembled lyotropic liquid crystals (LLCs) are a promising material for
  selective separations. To date, their separation performance has only been
  characterized with membranes made from a small set of LLC monomers. There
  remains a large design space that we can likely exploit to design membranes for
  solute specific separations. In this work, we develop an atomistic molecular
  model of an LLC membrane. We show that our model is maximally consistent with
  experimental observations by comparing simulated X-ray Diffraction (XRD)
  patterns and calculated ionic conductivity to their experimental counterparts.
  We explore, in depth, the composition and structure of the nanopores in order
  to give insights that are not easily accessible to experimentalists. With the
  clear picture of the nanoscopic structure of these membranes gained from this
  work, we will be able to use the model to understand the mechanisms of small
  charged and uncharged solute transport within the nanopores, and eventually we
  will be able to screen large sets of modified monomers with their separation
  properties in mind.

@@

  Reverse osmosis (RO) and nanofiltration (NF) are membrane-based separation
  processes widely used to create potable or reusable water by removing salts and
  other dissolved compounds from water sources. RO membranes are typically thin
  film composite (TFC) membranes with a porous support layer and a dense
  crosslinked polymer active layer which near indiscriminately separates all
  solutes. NF membranes have porous architectures with pores on the order of 1 nm
  in size which separate primarily based on solute size
  \cite{van_der_bruggen_review_2003}.

  RO and NF are finding applications beyond primarily water-based
  purifications. Recovery of valued products from complex aqueous and organic
  waste is of increasing interest. For example, flowback water (FW) produced
  during hydraulic fracturing is a complex wastewater full of potentially
  valuable dissolved organic compounds \cite{dischinger_application_2017}. There
  is increasing pressure to reuse hydraulic fracturing water rather than dispose
  of it in order to reduce social and envrionmental impacts as well as cost
  \cite{theodori_hydraulic_2014}. Rather than dispose of the wastestream
  generated in the recycling process, we can instead use highly selective
  membranes in order to successfully recover the useful compounds.

@@

  Selective separation by a semipermeable membrane barrier is governed by two
  steps: barrier entry and transport through the barrier. A solute's ability to
  enter a given membrane barrier is decided by its size, shape, charge, polarity
  and how those factors combine to interact with the membrane material. The same
  properties affect the rate at which a solute travels across the barrier. Since
  undesired solutes may still enter the membrane, their diffusion rates relative
  to desired solutes are important
  \cite{gin_polymerized_2008,wijmans_solution-diffusion_1995}. To achieve optimum
  selectivity, it is necessary to design membranes in a way that one can tune the
  relative diffusivities of desired and undesired solutes.

  Lyotropic liquid crystals (LLCs) are amphiphilic molecules that have the
  ability to self-assemble into porous nanostructures \cite{smith_ordered_1997}
  which can be crosslinked to create mechanically strong membrane films with
  pores on the order of 1 nm in diameter \cite{zhou_supported_2005}. Unlike
  commercial NF membranes, LLC membrane pores are uniform in size. Since LLC
  membranes lack a pore size distribution, they inherently exhibit high
  selectivity due to their strict molecular weight cut-off (MWCO).  Additionally,
  the LLCs chosen to make membranes for this application are salts.  The membrane
  gains a net surface charge when counterions from head groups that line the pore
  walls escape into the feed solution in order to balance the gradients of
  concentration and electric potential. This effect, which causes the membrane to
  repel incoming charged species, is called Donnan exclusion
  \cite{donnan_theory_1995}.

  The feasiblity of LLC membranes has been demonstrated using LLCs that form
  the type 1 bicontinuous cubic (Q\textsubscript{I})
  \cite{hatakeyama_water_2011,hatakeyama_nanoporous_2010,carter_glycerol-based_2012},
  and the inverted hexagonal (H\textsubscript{II}) \cite{zhou_supported_2005}
  phases (See Figure~\ref{fig:bcc_v_hII}). When separating organic solutes from
  NaCl, Q\textsubscript{I} phase membrane filtration experiments have shown
  selectivity 2-3 times higher than commercial RO and 6-12 times higher than
  commercial NF membranes \cite{dischinger_application_2017}.  When separating a
  series of various sized dyes, the H\textsubscript{II} phase membrane showed
  complete rejection of dyes bigger than 1.2 nm in size
  \cite{zhou_supported_2005}.

  We will focus our current efforts around the H\textsubscript{II} phase since
  its geometry should give optimal transport performance compared to the
  Q\textsubscript{I} phase. The H\textsubscript{II} phase forms at room
  temperature in the presence of c.a.~10 wt\% water and consists of hexagonally
  packed, straight, hydrophilic pore columns\cite{smith_ordered_1997}. In the
  absence of water, neat monomer will form the same hexagonal columnar structure
  which, in literature, has been referred to as the Col\textsubscript{h}
  thermotropic phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).
  These hexagonal columnar phases represent the ideal geometry for membrane
  transport. The most promising Q\textsubscript{I} phase for membrane
  applications forms at 70\degree C when monomer is mixed with c.a.~20 wt\%
  glycerol\cite{carter_glycerol-based_2012}. The Q\textsubscript{I} phase
  consists of a network of three dimensionally interconnected pores (See
  Figure~\ref{fig:bcc_phases}). This complex geometry effectively adds a degree
  of tortuosity to the membrane which prevents optimal through-plane transport.
  Despite the geometric differences between the Q\textsubscript{I} and
  H\textsubscript{II} phase, the chemical environment within the pores is
  expected to be similar, and either membrane should be highly selective.

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[trim=-2cm 0 0 0, clip, height=4.5cm]{Dibrpyr14.eps}
        \includegraphics[height=4.5cm]{bcc_phases.png}
        \caption{}\label{fig:bcc_phases}
  \end{subfigure}
  \hspace{1cm}
  \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[trim=0 -0.5cm 0 -0.5cm, clip, height=4.5cm]{NaGA3C11.eps}
        \includegraphics[height=4.5cm]{hexagonal_packing.png}
        \caption{}\label{fig:hII_phase}
  \end{subfigure}
  \caption{The choice of LLC monomer leads to different favorable liquid
        crystalline phases. The monomer shown in (a) forms the Q\textsubscript{I} phase
        in either the Ia3d (left) or Pn3m (right) space group. The surfaces shown
        represent the interface between the hydrophilic and hydrophobic regions of unit
        cell. As pictured, we assume that each region occupies an equal volume. The
        monomer shown in (b) will form the hexagonal columnar phase
        (H\textsubscript{II} in the presence of water or Col\textsubscript{h} when no
         water is present).}\label{fig:bcc_v_hII}
  \end{figure}

@@

  Research into the H\textsubscript{II} phase of polymerized LLC membranes has
  been revived in recent years. During early stages of exploration, hexagonal
  mesophases formed by Na-GA3C11 (Figure~\ref{fig:assembly}) could not be
  macroscopically aligned, resulting in low flux membranes, and no clear route
  towards scalable and economical filtration. For this reason, research efforts
  focused on the Q\textsubscript{I} phase, whose geometry does not require
  alignment. In 2014, Feng et al.~showed that one can align 
  Col\textsubscript{h} mesophases using a magnetic field with subsequent
  cross-linking to lock the structure in place \cite{feng_scalable_2014}. In
  2016, Feng et al. showed that one could obtain the same result using a second
  technique termed soft confinement \cite{feng_thin_2016}. Current efforts are
  focused on extending the method to the H\textsubscript{II} phase and 
  characterizing the performance of these newly aligned systems.

@@

  Our current understanding of LLC systems is not rich enough to be able to
  precisely design membranes for specific separations. Over the past 20 years,
  H\textsubscript{II}-phase LLC polymer membrane studies have been limited
  primarily to Na-GA3C11 with some characterization done after minor structural
  modifications. Resel et al.~varied the length of the monomer tails and the
  counterion used and observed its affect on pore spacing
  \cite{resel_structural_2000}. In a later study of rejection performance, it was
  shown that membranes formed by cross-linked Na-GA3C11 in the
  H\textsubscript{II} phase cannot separate solutes less than 1.2 nm in diameter
  because the pores are too large \cite{zhou_supported_2005}. We do not yet
  understand how to controllably reduce the effective pore size or how to tune
  the chemical environment in the nanopores of this or related materials for
  small molecule separations. The only source of predictive modeling for
  LLC systems have been macroscopic models that likely do not adequately describe
  transport at these length scales \cite{hatakeyama_water_2011}. It will be
  challenging to efficiently narrow down the large design space in a laboratory
  setting without a robust model.

@@

  %BJC: I think the first sentence of the following paragraph is
  %necessary to put somewhere, but maybe not in this spot exactly
  In order to appropriately model transport in these ordered, nanoporous
  organic systems, we must first gain a thorough understanding of the nanoscopic
  structure of LLC polymer membranes. Our approach to constructing a general
  model will follow the development of a model of the assembly formed by
  Na-GA3C11 since it has sufficient experimental characterization. We have also
  narrowed our scope to the development of a model of the Col\textsubscript{h}
  phase membrane. Compared to the H\textsubscript{II} phase, the
  Col\textsubscript{h} phase is a simpler starting point, due to the absence of
  water, and has detailed experimental wide-angle X-ray scattering (WAXS)
  patterns useful for reconstructing structural data. The two phases appear to
  share similar structural characteristics since the pore spacings in each system
  are in close agreement \cite{feng_thin_2016,resel_structural_2000}.

  Despite having structural data, there is still information which experiment
  cannot definitively answer. Our current level of understanding suggests that
  monomers partition into layers with head groups facing towards the center, then 
  stack on top of each other to create pores that pack together hexagonally.
  There are several key questions that we will investigate in order to
  expand this understanding:

@@

  While past work describes the Col\textsubscript{h} phase as dry
  ~\cite{feng_scalable_2014}, it has been suggested by experimentalists, in
  unpublished communications, that it is likely that neat monomer leaches small
  amounts of ambient water. Experimentally, achieving a hexagonal phase with a
  completely dry system has proven difficult. If neat monomer sits in ambient
  conditions, its color turns from transparent to slightly opaque and a hexagonal
  phase forms. Although we will not explore whether water is necessary for
  self-assembly, we hypothesize that the hydrogen bonding network formed by the
  water may play a role in structuring the pores and holding together the
  hexagonal phase. We can look at correlation functions and use simulated X-ray
  diffraction patterns to see if there is any meaningful structural difference
  between a ``dry'' and ``wet'' system.

>>>>

  \begin{abstract}
  %MRS5: Abstract can contain a little less introduction material. Suggest the following cuts  
  Nanostructured porous membranes made from the cross-linked hexagonal phase of
  self-assembled lyotropic liquid crystals (LLCs) are a promising material for
  selective separations. 
  %To date, their separation performance has only been
  %characterized with membranes made from a small set of LLC monomers. There
  %remains a large design space that we can likely exploit to design membranes for
  %solute specific separations. 
  In this work, we develop an atomistic molecular
  model of an LLC membrane. We show that our model is maximally consistent with
  experimental observations by comparing simulated X-ray Diffraction (XRD)
  patterns and calculated ionic conductivity to their experimental counterparts.
  We explore, in depth, the composition and structure of the nanopores in order
  to give insights that are not easily accessible to experimentalists. 
  %MRS5: suggest:
  %With the
  %clear picture of the nanoscopic structure of these membranes gained from this
  %work, we will be able to use the model to understand the mechanisms of small
  %charged and uncharged solute transport within the nanopores, and eventually we
  %will be able to screen large sets of modified monomers with their separation
  %properties in mind.  
  The clearer picture of the nanoscopic structure of these membranes provided in this study will enable better understanding of the mechanisms of small molecule transport within these nanopores.

@@

%MRS5: want to get people engaged from the beginning. The first
%paragraph should describe a problem to be solved. Then you can get
%into the motivating details (like background of the alternatives, and
%why we need something different).  I feel like this is an
%introduction into different types of membranes, though it should be
%an introduction to the problem that this paper takes steps towards
%solving.

  Reverse osmosis (RO) and nanofiltration (NF) are membrane-based separation
  processes widely used to create potable or reusable water by removing salts and
  other dissolved compounds from water sources. RO membranes are typically thin
  film composite (TFC) membranes with a porous support layer and a dense
  crosslinked polymer active layer which near indiscriminately 
%MRS5: 
%  Maybe ``impermeable to all solutes?'' or ``separates all solutes from water.''  As written, it's not stated what the solutes are separated from.
  separates all solutes. 
NF membranes have porous architectures with pores on the order of 1 nm
  in size which separate primarily based on solute size
  \cite{van_der_bruggen_review_2003}.

%MRS5: a more compelling thesis is that we need membranes that are more selective than RO and NF.  
  RO and NF are finding applications beyond primarily water-based
  purifications. Recovery of valued products from complex aqueous and organic
  waste is of increasing interest. For example, flowback water (FW) produced
  during hydraulic fracturing is a complex wastewater full of potentially
  valuable dissolved organic compounds \cite{dischinger_application_2017}. There
  is increasing pressure to reuse hydraulic fracturing water rather than dispose
  of it in order to reduce social and envrionmental impacts as well as cost
  \cite{theodori_hydraulic_2014}. Rather than dispose of the wastestream
  generated in the recycling process, we can instead use highly selective
  membranes in order to successfully recover the useful compounds.

@@

%MRS5: under solution-diffusion, then the free energy to partition into the membrane is important as well, correct?
  Selective separation by a semipermeable membrane barrier is governed by two
  steps: barrier entry and transport through the barrier. A solute's ability to
  enter a given membrane barrier is decided by its size, shape, charge, polarity
  and how those factors combine to interact with the membrane material. The same
  properties affect the rate at which a solute travels across the barrier. Since
  undesired solutes may still enter the membrane, their diffusion rates relative
  to desired solutes are important
  \cite{gin_polymerized_2008,wijmans_solution-diffusion_1995}. To achieve optimum
  selectivity, it is necessary to design membranes in a way that one can tune the
  relative diffusivities of desired and undesired solutes.

%MRS5: maybe a more compelling thesis would be that LLC might have the capabilility to serve as selective membranes?
  Lyotropic liquid crystals (LLCs) are amphiphilic molecules that have the
  ability to self-assemble into porous nanostructures \cite{smith_ordered_1997}
  which can be crosslinked to create mechanically strong membrane films with
  pores on the order of 1 nm in diameter \cite{zhou_supported_2005}. Unlike
  commercial NF membranes, LLC membrane pores are uniform in size. Since LLC
  membranes lack a pore size distribution, they inherently exhibit high
  selectivity due to their strict molecular weight cut-off (MWCO). 
%MRS5: is there a citation for the strict MWCO? 
%MRS5: suggested change.
 Additionally, the LLCs chosen to make membranes for this application are salts, and therefore lead to 
Donnan exclusion of ions in solution.
 The membrane
  gains a net surface charge when counterions from head groups that line the pore
  walls escape into the feed solution in order to balance the gradients of
  concentration and electric potential. 
%This effect, which causes the membrane to
%  repel incoming charged species, is called Donnan exclusion

  \cite{donnan_theory_1995}.

  The feasiblity of LLC membranes has been demonstrated using LLCs that form
  the type 1 bicontinuous cubic (Q\textsubscript{I})
  \cite{hatakeyama_water_2011,hatakeyama_nanoporous_2010,carter_glycerol-based_2012},
  and the inverted hexagonal (H\textsubscript{II}) \cite{zhou_supported_2005}
  phases (See Figure~\ref{fig:bcc_v_hII}). When separating organic solutes from
  NaCl, Q\textsubscript{I} phase membrane filtration experiments have shown
%MRS5: for dashes between two numbers, use double -- for the correct length in the file. 
  selectivity 2--3 times higher than commercial RO and 6--12 times higher than
  commercial NF membranes \cite{dischinger_application_2017}.  When separating a
  series of various sized dyes, the H\textsubscript{II} phase membrane showed
  complete rejection of dyes bigger than 1.2 nm in size
  \cite{zhou_supported_2005}.

%MRS5: be more explicit about current efforts. Also use the
%opportunity of a thesis statement to convince the readership of
%something.  what do you want to convince them?
  We will focus our current efforts around the H\textsubscript{II} phase since
  its geometry should give optimal transport performance compared to the
  Q\textsubscript{I} phase. 
%MRS5: citation on this above?
  The H\textsubscript{II} phase forms at room
  temperature in the presence of c.a.~10 wt\% water and consists of hexagonally
  packed, straight, hydrophilic pore columns\cite{smith_ordered_1997}. In the
  absence of water, neat monomer will form the same hexagonal columnar structure
  which, in literature, has been referred to as the Col\textsubscript{h}
  thermotropic phase\cite{feng_scalable_2014} (See Figure~\ref{fig:hII_phase}).
  These hexagonal columnar phases represent the ideal geometry for membrane
  transport. 
%MRS5: maybe less about Q_I since it's not in this paper?  The introduction should be an introduction to the paper.
The most promising Q\textsubscript{I} phase for membrane
  applications forms at 70\degree C when monomer is mixed with c.a.~20 wt\%
  glycerol\cite{carter_glycerol-based_2012}. The Q\textsubscript{I} phase
  consists of a network of three dimensionally interconnected pores (See
  Figure~\ref{fig:bcc_phases}). This complex geometry effectively adds a degree
  of tortuosity to the membrane which prevents optimal through-plane transport.
  Despite the geometric differences between the Q\textsubscript{I} and
  H\textsubscript{II} phase, the chemical environment within the pores is
  expected to be similar, and either membrane should be highly selective.

  \begin{figure}
  \centering
  \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[trim=-2cm 0 0 0, clip, height=4.5cm]{Dibrpyr14.eps}
        \includegraphics[height=4.5cm]{bcc_phases.png}
        \caption{}\label{fig:bcc_phases}
  \end{subfigure}
  \hspace{1cm}
  \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[trim=0 -0.5cm 0 -0.5cm, clip, height=4.5cm]{NaGA3C11.eps}
        \includegraphics[height=4.5cm]{hexagonal_packing.png}
        \caption{}\label{fig:hII_phase}
  \end{subfigure}
%MRS5: not sure Q should be in here.
  \caption{The choice of LLC monomer leads to different favorable liquid
        crystalline phases. The monomer shown in (a) forms the Q\textsubscript{I} phase
        in either the Ia3d (left) or Pn3m (right) space group. The surfaces shown
        represent the interface between the hydrophilic and hydrophobic regions of unit
        cell. As pictured, we assume that each region occupies an equal volume. The
        monomer shown in (b) will form the hexagonal columnar phase
        (H\textsubscript{II} in the presence of water or Col\textsubscript{h} when no
         water is present).}\label{fig:bcc_v_hII}
  \end{figure}

@@

%MRS5: not a strong thesis.  Why has it been revived?  The reasons for that would be a better thesis.
  Research into the H\textsubscript{II} phase of polymerized LLC membranes has
  been revived in recent years. During early stages of exploration, hexagonal
  mesophases formed by Na-GA3C11 (Figure~\ref{fig:assembly}) could not be
  macroscopically aligned, resulting in low flux membranes, and no clear route
  towards scalable and economical filtration. For this reason, research efforts
  focused on the Q\textsubscript{I} phase, whose geometry does not require
  alignment. In 2014, Feng et al.~showed that one can align 
  Col\textsubscript{h} mesophases using a magnetic field with subsequent
  cross-linking to lock the structure in place \cite{feng_scalable_2014}. In
  2016, Feng et al. showed that one could obtain the same result using a second
  technique termed soft confinement \cite{feng_thin_2016}. Current efforts are
  focused on extending the method to the H\textsubscript{II} phase and 
  characterizing the performance of these newly aligned systems.

@@

%MRS5: this is a better thesis. This sentence sets out a problem to the solved.
  Our current understanding of LLC systems is not rich enough to be able to
  precisely design membranes for specific separations. Over the past 20 years,
  H\textsubscript{II}-phase LLC polymer membrane studies have been limited
  primarily to Na-GA3C11 with some characterization done after minor structural
  modifications. Resel et al.~varied the length of the monomer tails and the
  counterion used and observed its affect on pore spacing
  \cite{resel_structural_2000}. In a later study of rejection performance, it was
  shown that membranes formed by cross-linked Na-GA3C11 in the
  H\textsubscript{II} phase cannot separate solutes less than 1.2 nm in diameter
  because the pores are too large \cite{zhou_supported_2005}. We do not yet
  understand how to controllably reduce the effective pore size or how to tune
  the chemical environment in the nanopores of this or related materials for
  small molecule separations. The only source of predictive modeling for
  LLC systems have been macroscopic models that likely do not adequately describe
  transport at these length scales \cite{hatakeyama_water_2011}. It will be
  challenging to efficiently narrow down the large design space in a laboratory
  setting without a robust model.

@@

  %BJC: I think the first sentence of the following paragraph is
  %necessary to put somewhere, but maybe not in this spot exactly
  In order to appropriately model transport in these ordered, nanoporous
  organic systems, we must first gain a thorough understanding of the nanoscopic
  structure of LLC polymer membranes. 
%MRS5: repeated use of model below, both constructing and developing?
Our approach to constructing a general
  model will follow the development of a model of the assembly formed by
  Na-GA3C11 since it has sufficient experimental characterization. 
%MRS5: maybe clean up/motivate abit more about why Col_h is worth starting with.
We have also
  narrowed our scope to the development of a model of the Col\textsubscript{h}
  phase membrane. Compared to the H\textsubscript{II} phase, the
  Col\textsubscript{h} phase is a simpler starting point, due to the absence of
  water, and has detailed experimental wide-angle X-ray scattering (WAXS)
  patterns useful for reconstructing structural data. The two phases appear to
  share similar structural characteristics since the pore spacings in each system
  are in close agreement \cite{feng_thin_2016,resel_structural_2000}.

  Despite having structural data, there is still information which experiment
  cannot definitively answer. 
%MRS5: does a level of understanding suggest, or just is the level of understanding those facts?  and are there citations for these? are they inferred by the existing x-ray data.
Our current level of understanding suggests that
  monomers partition into layers with head groups facing towards the center, then 
  stack on top of each other to create pores that pack together hexagonally.
  There are several key questions that we will investigate in order to
  expand this understanding:

@@

  While past work describes the Col\textsubscript{h} phase as dry~\cite{feng_scalable_2014}, 
  it has been suggested by experimentalists, in
  unpublished communications, that it is likely that neat monomer leaches small
  amounts of ambient water. Experimentally, achieving a hexagonal phase with a
  completely dry system has proven difficult. If neat monomer sits in ambient
  conditions, its color turns from transparent to slightly opaque and a hexagonal
  phase forms. Although we will not explore whether water is necessary for
  self-assembly, we hypothesize that the hydrogen bonding network formed by the
  water may play a role in structuring the pores and holding together the
  hexagonal phase. We can look at correlation functions and use simulated X-ray
  diffraction patterns to see if there is any meaningful structural difference
  between a ``dry'' and ``wet'' system.
